### 2020-10-15 doc/manuscript/Dow_manuscript.Rmd
edits

#relavant? reworded first line above#
(Understanding how growth is influenced by multiple factors at once is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation. CO2 concentrations in the atmosphere are increasing, and in a perfect scenario, trees and other plants would simply speed up growth in response to this, increasing photosynthetic efficiency in a process called CO2 fertilization (Strain and Cure, 1986; Huang et al., 2007). However, factors other than CO2 also limit tree growth, and seasonal temperature averages and precipitation patterns are changing concurrently. It is anticipated that, throughout much of the United States, future heat and drought may limit tree growth more strongly than it can be enhanced by elevated CO2 (Charney et al, 2016).) 
#

In recent decades, an increase in average Spring temperatures has been attributed to climate change. This relatively small increase has already been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming on leaf phenology have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018) leading to many interesting observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by earlier leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of Spring warming also has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

However, the fine-scale responses of temperate deciduous species to a warming Spring are not commonly researched due to the increased effort required. Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). Many observations of climate induced phenological shifts are based on GIS, eddy-covarience measurements, or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). To date, very little research, has been done on the phenology of stem-growth.

In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ). For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered contributing to a negative climate feedback loop (source?). Recent results from a paper by D’Orangeville et al. (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology. They showed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time. This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses (source for RP vs DP?).

While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

@@

Our first study site was the forest dynamics plot located at the Smithsonian COnservation Biology Institute 5 KM South of Front Royal, Virginia within the Blue Ridge Mountains (38 °53036.6″ N, 78°08043.4″W; elevation 273–338 m). The forest here was established in the mid-19th century after conversion from agricultural fields. Dominate canopy species within the 25.6 HA forest plot include tulip poplar (Liriodendron tulipifera), hickories (Carya spp.), and oaks (Quercus spp.). The climate is humid temperate, with a mean annual precipitation of 998 mm, January temperatures averaging 1 degrees C and July temperatures averaging 24 degrees C (Helcoski et al, 2019).  
The second study site was Harvard Forest in the central Massachusetts town of Petersham (42°53′ N, 72°17′ W, elevation?). Harvard forest was established in 1907 as a study site for university students and researchers. (Previous use?) Dominate species within the approximately 13 HA plot are hemlock (Tsuga spp.), oak (Quercus spp.) and maple (Acer spp.). The climate is temperate continental with a mean annual precipitation of 1120 mm, January temperatures averaging -12 degrees C and July temperatures averaging 19 degrees C (D'Orangeville et al, ).

@@

Within the plot, metal dendrometer bands were installed on *>#* trees and measured approximately every 1-2 weeks within the growing season for *9* years (*list ranges, for each site if different*). *(don't bother to mention biannual bands, unless you end up using the data--which you could, for annual growth, if you'd like to increase the sample size. You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census *(Table S1)*. Across years, the number of bands sampled averaged *~150* at SCBI, # at Harvard Forest*, and # at SERC*.   *(summary of how bands were distributed across species, focusing on ring/ diffuse porous distinction. details in an SI table)*

(After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees *--put this in Table S1*).

Measurements were timed to begin before beginning of spring growth and to continue through the cessation of growth in the fall. *(Give specifics, which may vary by site. Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)*

@@

### Discussion
-Growth timing
We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate(). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990). The pre-leaf stem growth occurring in these species is fueled by carbon stores, often used for other cell maintenance activities within the plant (). The dependence on these carbon stores for cambial activation could prove to be a detriment to the tree, as exhaustion of reserves leads to carbon starvation (). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer. This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.

The mechanisms controlling activation of cambium reactivation is still a mystery, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves(). However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.

Ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while 75% remains statistically unchanged, in response to warmer temperatures during early Spring. This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. The reactivation of cambial cells is likely beginning earlier potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves which may be in short demand as they are already being used for initial growth. Hypothetically, this could lead to carbon starvation in some trees. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts. In the long term, this could lead to changes in species dominance within temperate forests (talk more aobut this?).

Meanwhile, diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps. This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase. However, diffuse-porous trees are adapted to survive in drought conditions, and may continue to be successful even in future climate scenarios (RIght? seeing conflicting reports. Sources.). Studies have found that these diffuse-porous trees in the Eastern USA show the largest reduction in ring-width increment during, and following, Summer droughts (Kannenberg et al, 2018) as well as being generally more sensitive to Summer precipitation patterns (Elliot et al, 2015). The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition.  --the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades? Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)

This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is the metaphorical tip of the ice berg. Here, we focused our analysis purely on early Spring warming. This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth. For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?). This revelation agrees with the results from Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year using tree cores. Additionally, Other studies have found that Autumn leaf senescence is being delayed potentially in response to warmer temperatures in the previous year's Winter or Autumn (), changes in chilling requirements due to current year Autumn and winter temperatures (), and stunted total productivity induced by drought and high TMAX in the current year's Summer (), diffuse porous trees increased growth during the following growing season when winter temperatures (TMIN) were lower (Benito & Pederson, 2015; Elliot et al, 2015). How these relationships may impact intraannual growth patterns or variables like those described here is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

@@

Content to incorporate: N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact (Zohner et al., 2020)

>>>>

In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming are most apparent when examining leaf phenology shifts, which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many interesting observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of Spring warming also has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous species to a warming Spring are not commonly researched due to the increased effort required. Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). To date, very little research, has been done on the phenology of stem-growth.

In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ). For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered, contributing to a negative climate feedback loop (source?). Recent results from a paper by D’Orangeville et al. (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology. They observed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time. This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses.

^More info on wood-type? move some stuff up from discussion?

While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. However, traditional methods of direct observation are labor intensive, nad very time consuming (Source about cell sampling). A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

@@

Our first study site was the forest dynamics plot located at the Smithsonian COnservation Biology Institute 5 KM South of Front Royal, Virginia within the Blue Ridge Mountains (38 °53036.6″ N, 78°08043.4″W; elevation 273–338 m). The forest here was established in the mid-19th century after conversion from agricultural fields. Dominate canopy species within the 25.6 HA forest plot include tulip poplar (Liriodendron tulipifera), hickories (Carya spp.), and oaks (Quercus spp.). The climate is humid temperate, with a mean annual precipitation of 998 mm, January temperatures averaging 1 degrees C and July temperatures averaging 24 degrees C (Helcoski et al, 2019).

The second study site was Harvard Forest in the central Massachusetts town of Petersham (42°53′ N, 72°17′ W, elevation?). Harvard forest was established in 1907 as a study site for university students and researchers. (Previous use?) Dominate species within the approximately 13 HA plot are hemlock (Tsuga spp.), oak (Quercus spp.) and maple (Acer spp.). The climate is temperate continental with a mean annual precipitation of 1120 mm, January temperatures averaging -12 degrees C and July temperatures averaging 19 degrees C (D'Orangeville et al, ).

@@

Within the plot, metal dendrometer bands were installed on *>#* trees and measured with a digital caliper approximately every 1-2 weeks within the growing season for *9* years (*list ranges, for each site if different*). *(don't bother to mention biannual bands, unless you end up using the data--which you could, for annual growth, if you'd like to increase the sample size. You can put exact numbers of bands per year in an SI file.*)  The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census *(Table S1)*. Across years, the number of bands sampled averaged *~150* at SCBI, # at Harvard Forest*, and # at SERC*. Four main species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species.    *(summary of how bands were distributed across species, focusing on ring/ diffuse porous distinction. details in an SI table)*

(Bands After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees *--put this in Table S1*).

(Bands used after minimum growth checks: 
1  2011    84
2  2012    84
3  2013   100
4  2014   100
5  2015    96
6  2016    94
7  2017    89
8  2018    88
9  2019    75) this seems like alot are being thrown out - manually check if WAI
Measurements were timed to begin before beginning of spring growth and to continue through the cessation of growth in the fall. *(Give specifics, which may vary by site. Note that we adjusted timing at SCBI when we observed early leaf-out of understory shrubs/herbs)* The median start date was 4/14, which was adjusted when early leaf-out of understory shrubs and herbs was observed, and the median end date was 10/17, coinciding with late senesence in the plot. The earliest start date for intraannual measurements was in 2020 on March 30, and the latest end date was on 2012 on November 26.

@@

### Discussion

We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994). The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014). The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994). This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.

The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005). However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production.

Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth remains statistically unchanged. This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. In fact, we found a positive correlation between growing season length in ring-porous trees and Spring temperature, while no such relationship was observed in diffuse-porous trees. The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.

Diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps. This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas. ---- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015). The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.

--the range of some of these species may migrate North to cooler summer conditions becoming potentially extant in the Southern / mid Atlantic of the USA.--    (Is mortality of ring-porous species higher in recent decades? Hard to pinpoint, but would likely be caused by warmer early-spring followed by late frost events exhausting NSC reserves?)

This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg. Here, we focused our analysis purely on early Spring warming. This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth. For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?). This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year. Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high TMAX in the current year's Summer (), and an increase in diffuse-porous tree growth during the gorwing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015). How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

@@

Content to incorporate: N. American strategies have conservative strategies when it comes to phenology, as historically they’ve been subject to more spring frosts. Thus, climate change is having less impact (Zohner et al., 2020); Leaf velocity discussion?
